Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-03-19 manuscriptPGPEsuppl.tex

\begin{figure}[]
  \centering
  \includegraphics[width=12.0cm]{./Figs/DIHEDRALSlipid17eccWITHCaClPOPC.eps}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G1_M_M_G2_M_M_G3_M_M_G3O1_Mlipid17ecc.png}  
  \caption{\label{DIHSwithCAlipid17eccPOPC}
    Changes in POPC lipid17ecc dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
   \includegraphics[width=12.0cm]{./Figs/DIHEDRALScharmm36WITHCaClPOPC.eps}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
  \caption{\label{DIHSwithCAcharmm36POPC}
    Changes in POPC CHARMM36 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MSlipids.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MSlipids.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MSlipids.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MSlipids.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MSlipids.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MSlipids.png}  
%  \caption{\label{DIHSwithCA}
%    Changes in POPC Slipids dihedrals with increasing amount of CaCl$_2$.
%  }
%\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSslipidsWITHCaClPOPG.eps}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3C6_MSlipids.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MSlipids.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MSlipids.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MSlipids.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MSlipids.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MSlipids.png}  
  \caption{\label{DIHSwithCAslipidsPOPG}
    Changes in POPG Slipids dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_Mlipid17.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_Mlipid17.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_Mlipid17.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_Mlipid17.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_Mlipid17.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPC_M_G1_M_M_G2_M_M_G3_M_M_G3O1_Mlipid17.png}  
%  \caption{\label{DIHSwithCA}
%    Changes in POPC lipid17 dihedrals with increasing amount of CaCl$_2$.
%  }
%\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSlipid17WITHCaClPOPG.eps}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3C6_Mlipid17.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_Mlipid17.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_Mlipid17.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_Mlipid17.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_Mlipid17.png}
  %\includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G1_M_M_G2_M_M_G3_M_M_G3O1_Mlipid17.png}  
  \caption{\label{DIHSwithCAlipid17POPG}
    Changes in POPG lipid17 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3C6_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_Mlipid17ecc.png}
%  \includegraphics[width=6.0cm]{./Figs/dihedralFIGS/POPG_M_G1_M_M_G2_M_M_G3_M_M_G3O1_Mlipid17ecc.png}  
%  \caption{\label{DIHSwithCA}
%    Changes in POPG lipid17ecc dihedrals with increasing amount of CaCl$_2$.
%  }
%\end{figure}

@@

The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).
The large set of MD simulation data was analysed using the development version of NMRlipids databank.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using Python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations are available in Ref.~\citenum{MATCHgit}
(\texttt{scripts/calcOrderParameters.py}). For united atom simulations, the trajectories with hydrogens having ideal geometry were constructed first using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories. This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
The number density profiles of ions were calculated using the \texttt{gmx density} tool of the GROMACS sofware package \cite{gromacsMANUAL}.

@@

\noindent {\it POPG} \todo{Simulation details by Ollila}.

@@

\noindent {\it POPC:POPG mixture with additional calcium} \todo{Simulation details by A. Kiirikki.}

@@

\noindent {\it POPC:POPG} \todo{Simulation details by S. Virtanen or O. H. S. Ollila.}

>>>>

\begin{figure}[]
  \centering
  \includegraphics[width=12.0cm]{./Figs/DIHEDRALSlipid17eccWITHCaClPOPC.eps}
  \caption{\label{DIHSwithCAlipid17eccPOPC}
    Changes in POPC lipid17ecc dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
   \includegraphics[width=12.0cm]{./Figs/DIHEDRALScharmm36WITHCaClPOPC.eps}
  \caption{\label{DIHSwithCAcharmm36POPC}
    Changes in POPC CHARMM36 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSslipidsWITHCaClPOPG.eps}
  \caption{\label{DIHSwithCAslipidsPOPG}
    Changes in POPG Slipids dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=16.0cm]{./Figs/DIHEDRALSlipid17WITHCaClPOPG.eps}
  \caption{\label{DIHSwithCAlipid17POPG}
    Changes in POPG lipid17 dihedrals with increasing amount of CaCl$_2$.
  }
\end{figure}

@@

The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
The simulations were analyzed using preliminary versions of the NMRlipids databank
(\url{www.nmrlipids.fi}, \url{github.com/NMRlipids/MATCH} and \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/tree/master/Data/Simulations})
and unique naming convention for lipid atoms (\url{http://nmrlipids.blogspot.com/2015/03/mapping-scheme-for-lipid-atom-names-for.html}),
which enable automatic analysis of simulations with different force fields with varying atom naming conventions.
The automatic analyses were implemented using MDAnalysis \cite{agrawal11,gowers16} and MDTraj \cite{mcgibbon15}
python libraries, and tools in the GROMACS sofware package \cite{gromacsMANUAL}.
All codes are available from the project's GitHub repository \cite{NMRlipidsIVbgit}.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Code for all atom simulations is available in Ref.~\citenum{MATCHgit} (\texttt{scripts/calcOrderParameters.py}).
For united atom simulations, we first constructed trajectories including hydrogens with ideal geometry using either \texttt{buildH} program~\cite{buildH} or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were then calculated from these trajectories. This approach has been tested against trajectories with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\

@@

\noindent {\it POPG} Lipid bilayer containing 118 POPG molecules, 4110 TIP3P water molecules, and 118 potassium ions was build using CHARMM-GUI \cite{lee16}.
The system was simulated 100~ns, coupled to 298 K using Nose-Hoover~\cite{nose84,hoover85} thermostat and 1 bar with
semi-isotropic Parrinello-Rahman~\cite{parrinello81} pressure coupling.
The used default parameters and force field files from CHARMM-GUI are available from \citenum{CHARMM36popg}.

\todo{Simulation details for larger simulation by A. Peon.}

@@

\noindent {\it POPC:POPG 1:1 and POPC:POPG 4:1 mixtures with additional calcium} 
The initial structures were built with CHARMM-GUI Membrane Builder \cite{lee16}. The TIP3P water model was used to solvate the systems. The simulations were run for 400~ns with timestep 2~fs and the first 100~ns were discarded as equilibration time. The simulations were run with GROMACS version 2020.2~\cite{pall20}. The Nose-Hoover thermostat~\cite{nose84,hoover85} was used with temperature of 298~K and the time constant for temperature coupling was 1.0~ps. The semi-isotropic Parinello-Rahman barostat~\cite{parrinello81} was used with reference pressure 1.0~bar and with a time constant of 5.0~ps with compressibility of 4.5e-5~bar$^{-1}$. Long range electrostatic interactions were calculated with the PME method. All bonds with hydrogen atoms were constrained with LINCS algorithm.
The simulation files are available from Refs.~\citenum{CHARMM36POPCPOPG5050,CHARMM36POPCPOPG5050100mMCaCl,CHARMM36POPCPOPG50501000mMCaCl,CHARMM36POPCPOPG4010,CHARMM36POPCPOPG4010100mMCaCl,CHARMM36POPCPOPG40101000mMCaCl}.

@@

\noindent {\it POPC:POPG 4:1 and POPC:POPG 1:1 mixtures}
Initial structures were build by removing appropriate amount of lipids from POPC:POPG 7:3 mixture available at Ref.~\cite{POPCPOPG73lipid17}.
Force field parameters from the same reference were used \todo{We still need description from A. Peon how these were obtained},
except that incorrect dihedrals with type 1 were changed to type 9
(for details, see discussion in here \url{https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/12}).
Simulations were performed using the Gromacs simulation package~\cite{pall20} with time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff; the Verlet scheme was used\cite{Pall13};
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS~\cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps.
